---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
options(width = 80)
```

# sessioncheck

<!-- badges: start -->
[![R-CMD-check](https://github.com/djnavarro/sessioncheck/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/djnavarro/sessioncheck/actions/workflows/R-CMD-check.yaml)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![Codecov test coverage](https://codecov.io/gh/djnavarro/sessioncheck/graph/badge.svg)](https://app.codecov.io/gh/djnavarro/sessioncheck)
<!-- badges: end -->

The goal of **sessioncheck** is to provide simple tools that can be called at the top of a script, and produce warnings or errors if it detects signs that the script is not being executed in a clean R session. The intended user for **sessioncheck** is a beginner or intermediate level R user who needs a drop-in replacement for the simple but unsafe method of calling `rm(list = ls())` at the top of the script.

## Installation

You can install the development version of sessioncheck from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("djnavarro/sessioncheck")
```

## Example

The intention when using **sessioncheck** is that you would rarely load it with `library()`. Instead, a single line of code like this would be added at the top of the script:

```{r}
#| label: default-behaviour
sessioncheck::sessioncheck()
```

The default behaviour is to check for objects in the global environment and to check packages and environments attached to the search path and produce a warning if issues are detected. This can be converted to an error if a stricter check is required, and additional checks can be added if desired. For details on how the checks work and how they can be customised see the [get started](https://sessioncheck.djnavarro.net/articles/sessioncheck.html) page.
