---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# sessioncheck

<!-- badges: start -->
[![R-CMD-check](https://github.com/djnavarro/sessioncheck/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/djnavarro/sessioncheck/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of **sessioncheck** is to provide simple tools that can be called at the top of a script, and produce warnings or errors if it detects hints that it is not being executed in a clean R session. It is not intended as a replacement for sophisticated tools such as **targets**, **renv**, **callr**, and the like. These are tools that expert R users can use to impose tight controls over how a script is executed, and for expert-level users these are strongly recommended over **sessioncheck**. The purpose of **sessioncheck** is to provide a collection of simple tools that novice or intermediate level R users can use, as a drop-in replacement for the common (but unsafe) approach of placing `rm(list = ls())` at the top of the script.

## Installation

You can install the development version of sessioncheck from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("djnavarro/sessioncheck")
```

## Example

```{r example}
# Basic use
sessioncheck::check_environment()
sessioncheck::check_packages()

# To inspect the detected packages, assign the results
# to a variable. The pkg variable is a logical vector,
# with names corresponding to the detected packages. 
# Values are set to TRUE for base packages as well as 
# any packages included in the ignore list
pkg <- sessioncheck::check_packages()
pkg


# The same applies to objects detected in the global
# environment. Default behaviour is to permit hidden
# variables (i.e., those with names that start with 
# a dot)
obj <- sessioncheck::check_environment()
obj
```

