---
title: "sessioncheck"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{sessioncheck}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The goal of **sessioncheck** is to provide simple tools that can be called at the top of a script, and produce warnings or errors if it detects signs that the script is not being executed in a clean R session. 

## Why is this important?

A common practice when writing R scripts is to include a snippet of code like `rm(list = ls())` at the top of the script. The reason people do this is for reproducibility purposes, to ensure that the script is run in the context of a "clean" R session. The goal is a good one, but the method used to attain it is not very effective. The only thing that `rm(list = ls())` does is remove any variables currently stored in the global environment, but this is only one of many different ways in which previously-executed code can alter the state of the R session. This approach doesn't provide any degree of safety from any of the following:

- packages may have been loaded with `library()`
- data sets may have been added to the search path with `attach()`
- options may have been set with `options()`
- "hidden" variables in the global environment like `.Random.seed`
- (...and a long list of other more subtle things)

Any of these can affect how your script executes, and `rm(list = ls())` does not protect you against any of them. Because of this, a better practice is to **restart the R session** immediately before running the script. By running the script in a fresh R session, you're much less likely to encounter these issues.

## How does sessioncheck help?

The principle that underpins **sessioncheck** is that scripts should not attempt a "clean up" of the environment. A better approach is to include a snippet of code that inspects the state of the R session, and produces a warning (or even an error) if it detects signs that the script is not being run in a clean session. The simplest way to do this is with the `check_session()` function. The user could put this as the very line of their R script:

```{r}
#| label: basic-use
#| eval: false
sessioncheck::check_session("error")
```

When executed in a clean environment this function does nothing, and the script will continue to run as it normally would. However, if the session is "dirty" -- for some specific interpretation of what that means -- it will produce an error, prevent the script from running, and prompt the user to consider restarting the R session. 

This intended behaviour is very easy to illustrate in this vignette, because the environment inside the R markdown document has been altered in many ways that make it different from a "typical" clean R session. Here is an example showing the error that a user might see if they run a "session checked" script in a dirty R session:

```{r}
#| label: error-example
#| error: true
sessioncheck::check_session("error")
```

The error message lists two specific problems:

- several "unapproved" packages have been loaded via namespace
- one "unapproved" environment has been appended to the search path

This is then followed by a prompt for the user to restart R.

## How does sessioncheck work?

```{r}
#| label: individual-checks
library(sessioncheck)
check_globalenv()
check_packages()
check_attachments()
check_namespaces()
```
