---
title: "Individual session check functions"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The **sessioncheck** package is built on several functions that each check one specific aspect to the R session: the `sessioncheck()` function itself merely aggregates the results of individual check functions like `check_globalenv()`, `check_packages()` and `check_attachments()`. In a typical workflow the user would be unlikely to use these directly, relying instead on the high-level interface provided by `sessioncheck()`. Nevertheless, it's useful to look at what each one does in a little more detail. 

### Check the global environment

The first and simplest of the checks is `check_globalenv_objects()`, and it focuses on the same aspect of the R session that the traditional `rm(list=ls())` method does: the contents of the global environment. At the moment there is nothing in the global environment, so it is considered "clean". As a consequence, nothing happens when we run this check:

```{r}
#| label: check-global-environment-1
sessioncheck::check_globalenv_objects()
```

To get the check to produce a warning, we'll need to add some variables:

```{r}
#| label: check-global-environment-2
visible_1 <- "this will get detected"
visible_2 <- "so will this"
.hidden_1 <- "but this will not"

sessioncheck::check_globalenv_objects()
```

The output indicates that the script has detected `visible_1` and `visible_2` in the global environment, and issues a warning to suggest that the R session may be contaminated.

There are two arguments to `check_globalenv_objects()`:

- `action` specifies what the function should do if an issue is detected. There are four allowed values: `error`, `warn` (the default), `message`, and `none`.
- `allow_globalenv_objects` is a character vector used to specify the rules that are used to decide *which* objects should trigger an action. A variable name that is included in the `allow` list will not trigger an action. There is a special case: `allow_globalenv_objects = NULL` will apply the same rule that `ls()` uses when listing the contents of the global environment: variables that start with a `.` will be ignored, and will not trigger an action. 

The example below illustrates how both of these actions are used. Here, the action taken will be to print a message rather than a warning; and by setting the `allow_globalenv_objects` argument to an empty string, *any* variable in the global environment will trigger the message, even the "hidden" ones:

```{r}
#| label: check-global-environment-3
sessioncheck::check_globalenv_objects(action = "message", allow_globalenv_objects = "")
```

This time we notice that the check detects the `visible_1` and `visible_2` like last time, but it now detects two hidden variables: the `.hidden_1` variable that I created earlier, and also the `.Random.seed` variable that R uses to store the state of the random number generator.

### Check the attached packages

```{r}
#| label: check-packages-1
sessioncheck::check_attached_packages()
```

The warning notes that the **sessioncheck** package has been attached. This might be considered acceptable, so we can ask the check to `allow` this package:

```{r}
#| label: check-packages-2
sessioncheck::check_attached_packages(action = "warn", allow_attached_packages = "sessioncheck")
```

### Check other attached environments

```{r}
#| label: check-attached-environments-1
sessioncheck::check_attached_environments()
```


### Check loaded namespaces

```{r}
#| label: check-loaded-namespaces-1
sessioncheck::check_loaded_namespaces()
```

### Check session runtime

```{r}
#| label: check-loaded-sessiontime-1
sessioncheck::check_sessiontime()
```

### Check required options

```{r}
#| label: check-required-options-1
sessioncheck::check_required_options()
```

### Check required locale

```{r}
#| label: check-required-locale-1
sessioncheck::check_required_locale()
```

### Check required environment variables

```{r}
#| label: check-required-sysenv-1
sessioncheck::check_required_sysenv()
```


