---
title: "S3 classes and methods"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(sessioncheck)
```

The **sessioncheck** package is designed to be lightweight, and has no dependencies. It defines three S3 classes: the individual check functions return objects of class `sessioncheck_status`, the `sessioncheck()` function returns an object of class `sessioncheck_sessioncheck`, and `sessionstate()` returns an object of class `sessioncheck_sessionstate`.

```{r}
pkg_status <- check_attached_packages(action = "none")
obj_status <- check_globalenv_objects(action = "none") 
session_check <- sessioncheck(action = "none")
session_state <- sessionstate()

class(pkg_status)
class(obj_status)
class(session_check)
class(session_state)
```

All three classes have a `format()` method and a `print()` method, and all three are coercible to data frames via `as.data.frame()`. A brief discussion of these classes and methods is provided here.

## sessioncheck_status objects

Internally, a `sessioncheck_status` object is a named list with two components:

- The `status` component is a named logical vector where names correspond to detected entities, and values that indicate whether an action is triggered by that entity. It is intended to be roughly analogous to an "exit status": `FALSE` corresponds to exit status zero and no actions are triggered, `TRUE` is a non-zero exist status and actions can be taken.
- The `type` component is a string indicating which check function created the status object.

This is illustrated below:
```{r}
unclass(pkg_status)
```

The `format()` and `print()` methods for a `sessioncheck_status` object produce the text displayed to the user in any message, warning, or error:

```{r}
print(pkg_status)
```

As a convenience, an `as.data.frame()` method is also supplied:

```{r}
as.data.frame(pkg_status)
```

## sessioncheck_sessioncheck objects

Because the `sessioncheck()` function can call multiple checks, the data structure it returns is slightly different. As illustrated below it is a named list of `sessioncheck_status` objects:

```{r}
lapply(session_check, unclass)
```

As before, the `format()` and `print()` methods are used to construct the text to be displayed to the user:

```{r}
print(session_check)
```

Similarly, there is an `as.data.frame()` method provided as a convenience:

```{r}
as.data.frame(session_check)
```

## sessioncheck_sessionstate objects

Unlike the two classes above, a `sessioncheck_sessionstate` object isn't a check result -- it's a snapshot of session state produced by `sessionstate()`, intended as an audit record rather than a pass/fail judgment. Internally it's a named list with twelve components (`platform`, `locale`, `matrix`, `document`, `machine`, `git`, `timing`, `rng`, `libpaths`, `packages`, `globalenv`, `attachments`), most of which are themselves lists or data frames rather than simple vectors:

```{r}
str(unclass(session_state), max.level = 1)
```

Because there's a lot captured here, the `format()` and `print()` methods accept optional arguments (one per component, plus `globalenv_n`) that select which fields or columns to display, without touching the underlying object:

```{r}
print(session_state, packages = "package", globalenv = "class", machine = character(0))
```

See `?display_methods` for the complete list of these arguments and how their defaults are resolved.

The `as.data.frame()` method behaves differently here than it does for the other two classes: rather than coercing every component into one combined table, it returns just the `packages` data frame, since that's the one component of `sessionstate()` that was already tabular and is the most natural fit for `as.data.frame()`.

```{r}
head(as.data.frame(session_state))
```

The other components remain available via `session_state$globalenv`, `session_state$attachments`, and so on -- see [sessionstate() versus sessionInfo() and session_info()](sessionstate-vs-sessioninfo.html) for a discussion of what each component captures.
