---
title: "Reporting the session state"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(sessioncheck)
```

In addition to tools for checking the state of an R session, **sessioncheck** also provides a `sessionstate()` function that can be used to report on the overall state of an R session. This function is similar in spirit to the [`utils::sessionInfo()`](https://rdrr.io/r/utils/sessionInfo.html) function available in base R, and also to the [**sessioninfo**](https://sessioninfo.r-lib.org/) package that offers a more detailed replacement, `sessioninfo::session_info()`. This article explains what `sessionstate()` does, and the ways in which it is similar to but also different from the existing tools. In particular, it discusses what `sessionstate()` adds over and above what the other two functions report, and -- because it does add quite a lot of new information -- what you give up in return, namely some privacy.

## Comparing two snapshots

A single `sessionstate()` call is a snapshot of one moment; on its own it doesn't say what changed. `compare_sessionstates()` takes two snapshots and reports how they differ -- take one at the start of a script (or an interactive session), do some work, take another, and diff them:

```{r}
# wrapped in a function so `baseline`/`current` themselves never end up in
# .GlobalEnv -- otherwise each would show up as "added" in the diff below,
# simply for having been assigned in between the two snapshots
take_diff <- function() {
  baseline <- sessionstate()
  assign("some_result", 1:10, envir = .GlobalEnv)
  current <- sessionstate()
  compare_sessionstates(baseline, current)
}
diff <- take_diff()
diff
```

```{r, include = FALSE}
rm(some_result, envir = .GlobalEnv)
```

By default, `print()`/`format()` collapse every section with nothing to report down to a single "(no changes)" line (`changed_only = TRUE`, as seen above for every section but `globalenv` and `timing`); pass `changed_only = FALSE` to see every field for the record-shaped sections (`platform`, `locale`, `matrix`, `document`, `machine`, `git`, `rng`) regardless of whether it changed. `timing` is always shown in full, since `captured_at`/`elapsed_sec` necessarily differ between any two `sessionstate()` calls -- there's no "unchanged" case to collapse.

Like `sessionstate()`'s three tabular sections, the diff's `packages`/`globalenv`/`attachments` sections can be coerced with `as.data.frame()`, though the shape is different: one row per key and tracked field, tagged `"added"`, `"removed"`, or `"modified"` rather than one row per key overall.

```{r}
as.data.frame(diff, which = "globalenv")
```

`globalenv`'s rows also carry a `verified` column. `sessionstate()` fingerprints each global environment object's value (an MD5 hash of its serialized form), so a value change can be detected even when an object's class and size stay the same; when both snapshots have a usable hash for an object, a hash mismatch is authoritative (`verified = TRUE`). When an object can't be hashed at all (e.g. one holding a database connection), the comparison falls back to `class`/`size` only (`verified = FALSE`) -- and if neither of those changed either, a value change could have happened invisibly.

`compare_sessionstates()` also warns if `new` looks like it was captured *before* `old`, in case the two arguments were passed in the wrong order. See `?compare_sessionstates` for the full per-section diff semantics.
