---
title: "Comparing session states"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The [session state reporting](sessionstate-reporting.html) article introduces `sessionstate()`, which takes a snapshot of the current R session -- packages, global environment contents, RNG state, and more. A single snapshot is useful as a standalone record, but on its own it can't answer a very natural follow-up question: *what changed?*

```{r setup}
library(sessioncheck)
```


## A concrete "what changed" question

Suppose you're debugging a script that behaves inconsistently: some runs produce the expected result, others don't, and you suspect the script itself is fine but something about the session it runs in differs from one run to the next. Or perhaps you're just cautious about a chunk of code you're about to run, and want to know afterwards exactly what it did to the session -- did it leave new objects behind? Change a package version? Consume random numbers you didn't expect it to?

Both questions have the same shape: take a snapshot before, take another after, and compare the two. `compare_sessionstates()` does the comparing.

## A basic comparison

```{r}
# wrapped in a function so `baseline`/`current` themselves never end up in
# .GlobalEnv -- otherwise each would show up as "added" in the diff below,
# simply for having been assigned in between the two snapshots
take_diff <- function() {
  baseline <- sessionstate()
  assign("some_result", 1:10, envir = .GlobalEnv)
  current <- sessionstate()
  compare_sessionstates(baseline, current)
}
diff <- take_diff()
diff
```

```{r, include = FALSE}
rm(some_result, envir = .GlobalEnv)
```

The comment in the previous code chunk is worth discussing further, because it highlights a real gotcha: `compare_sessionstates()` compares whatever is in `.GlobalEnv` at each snapshot, so any variable you assign there between the two calls -- including the baseline snapshot itself -- will show up as "added". Wrapping the two snapshots in a function keeps `baseline` and `current` local to that function, so the only real change visible in the diff is the one this example is trying to demonstrate: `some_result` being created.

## Reading the output

Most sections above collapse to a single "(no changes)" line -- that's `print()`'s default behavior (`changed_only = TRUE`), which hides any section where nothing differed between the two snapshots. `globalenv` is the exception here because `some_result` really was added; `timing` is always shown in full, since `captured_at`/`elapsed_sec` necessarily differ between any two `sessionstate()` calls, so there's no "unchanged" case for a diff to collapse. To see every field for the record-shaped sections (`platform`, `locale`, `matrix`, `document`, `machine`, `git`, `rng`) regardless of whether it changed, pass `changed_only = FALSE`.

## Digging into a specific section

Like `sessionstate()`'s tabular sections, `packages`/`globalenv`/`attachments` can be coerced with `as.data.frame()` for programmatic use. The shape is a little different from `sessionstate()`'s own tables, though: one row per key *and* tracked field, tagged `"added"`, `"removed"`, or `"modified"`, rather than one row per key overall.

```{r}
as.data.frame(diff, which = "globalenv")
```

## How much can you trust a "modified" value?

An important thing to pay attention to, when comparing two snapshots, is that there are limits to how far you can trust a "modified" value when it appears in the diff. Under the hood, `sessionstate()` creates a hash from each object stored in the global environment, so `compare_sessionstates()` can usually detect that a value has changed between two snapshots. When both snapshots have a usable fingerprint for an object, a mismatch is considered authoritative, and the comparison is marked `verified = TRUE` within the diff object itself.

However, this capability is limited in what it can accomplish. Some objects can't be fingerprinted at all -- one holding a live database connection, for example. When that happens, the only thing that `compare_sessionstates()` can rely on is the more generic `class` and `size` only. In those cases, the corresponding entry in the diff object is marked `verified = FALSE`: real changes could have occurred between snapshots, in a manner that the package is unable to check. In short, when you see `verified = FALSE` it should be treated as a signal to double-check manually. It doesn't mean that something has gone wrong: it is an acknowledgement that `compare_sessionstates()` was not able to make the comparison that we might wish it were able to.

Along the same lines, you should also be aware that the reverse pattern can happen for R6 objects, closures, and other environment-backed values. Because environments have different semantics to other kinds of R objects, it is possible that `verified = TRUE` can still report a spurious "modified" row, since fingerprinting is sensitive to binding order, not just content.

In short, caution is always advisable when interpreting the part of the output that describes how objects differ across two snapshots. It is a *heuristic* -- it is not and cannot be a definitive assertion about how two R session states differ. See `?compare_sessionstates` for the specifics of both failure modes.

## Pay attention to timing information

One final thing is worth mentioning: `compare_sessionstates()` warns if `new` looks like it was captured *before* `old`, since that usually means the two snapshots were passed in the wrong order. Here's what that looks like:

```{r}
show_backwards_order <- function() {
  early <- sessionstate()
  Sys.sleep(0.05)
  late <- sessionstate()
  invisible(compare_sessionstates(late, early))
}
show_backwards_order()
```

The comparison is still computed either way -- this is a nudge to double-check your arguments, not an error. See `?compare_sessionstates` for the complete per-section diff semantics, including how each of `sessionstate()`'s twelve elements is compared.
