---
title: "Comparing session states"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(sessioncheck)
```

The [session state reporting](sessionstate-reporting.html) article introduces `sessionstate()`, which takes a snapshot of the current R session -- packages, global environment contents, RNG state, and more. A single snapshot is useful as a standalone record, but on its own it can't answer a very natural follow-up question: *what changed?*

## A concrete "what changed" question

Suppose you're debugging a script that behaves inconsistently: some runs produce the expected result, others don't, and you suspect the script itself is fine but something about the session it runs in differs from one run to the next. Or perhaps you're just cautious about a chunk of code you're about to run, and want to know afterwards exactly what it did to the session -- did it leave new objects behind? Change a package version? Consume random numbers you didn't expect it to?

Both questions have the same shape: take a snapshot before, take another after, and compare the two. `compare_sessionstates()` does the comparing.

## A basic comparison

```{r}
# wrapped in a function so `baseline`/`current` themselves never end up in
# .GlobalEnv -- otherwise each would show up as "added" in the diff below,
# simply for having been assigned in between the two snapshots
take_diff <- function() {
  baseline <- sessionstate()
  assign("some_result", 1:10, envir = .GlobalEnv)
  current <- sessionstate()
  compare_sessionstates(baseline, current)
}
diff <- take_diff()
diff
```

```{r, include = FALSE}
rm(some_result, envir = .GlobalEnv)
```

That comment is worth pausing on, because it's a real gotcha rather than defensive over-caution: `compare_sessionstates()` compares whatever is in `.GlobalEnv` at each snapshot, so any variable you assign there between the two calls -- including `baseline` itself, if you didn't wrap things in a function -- will show up as "added". Wrapping the two snapshots in a function keeps `baseline` and `current` local to that function, so the only real change visible in the diff is the one this example is trying to demonstrate: `some_result` being created.

## Reading the output

Most sections above collapse to a single "(no changes)" line -- that's `print()`'s default behavior (`changed_only = TRUE`), which hides any section where nothing differed between the two snapshots. `globalenv` is the exception here because `some_result` really was added; `timing` is always shown in full, since `captured_at`/`elapsed_sec` necessarily differ between any two `sessionstate()` calls, so there's no "unchanged" case for a diff to collapse. To see every field for the record-shaped sections (`platform`, `locale`, `matrix`, `document`, `machine`, `git`, `rng`) regardless of whether it changed, pass `changed_only = FALSE`.

## Digging into a specific section

Like `sessionstate()`'s tabular sections, `packages`/`globalenv`/`attachments` can be coerced with `as.data.frame()` for programmatic use. The shape is a little different from `sessionstate()`'s own tables, though: one row per key *and* tracked field, tagged `"added"`, `"removed"`, or `"modified"`, rather than one row per key overall.

```{r}
as.data.frame(diff, which = "globalenv")
```

## How much can you trust a "modified" value?

`globalenv`'s rows carry a `verified` column, and it's worth knowing what it does and doesn't promise. `sessionstate()` fingerprints each global environment object's serialized value, so `compare_sessionstates()` can usually detect that a value changed even when its class and size stayed the same -- for instance, an object mutated in place. When both snapshots have a usable fingerprint for an object, a mismatch is authoritative, and the row is marked `verified = TRUE`.

Some objects can't be fingerprinted at all -- one holding a live database connection, for example. When that happens, the comparison falls back to `class`/`size` only, and the row is marked `verified = FALSE`: if neither of those changed either, a real value change could still have happened without being detected. `verified = FALSE` is a signal to double-check manually, not a sign that anything is broken.

## One thing worth watching for

`compare_sessionstates()` warns if `new` looks like it was captured *before* `old`, since that usually means the two snapshots were passed in the wrong order:

```{r}
show_backwards_order <- function() {
  early <- sessionstate()
  Sys.sleep(0.05)
  late <- sessionstate()
  invisible(compare_sessionstates(late, early))
}
show_backwards_order()
```

The comparison is still computed either way -- this is a nudge to double-check your arguments, not an error. See `?compare_sessionstates` for the complete per-section diff semantics, including how each of `sessionstate()`'s twelve elements is compared.
