---
title: "Why is it important to check the R session?"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

A common practice when writing R scripts is to include a snippet of code like `rm(list = ls())` at the top of the script. The reason people do this is for reproducibility purposes, to ensure that the script is run in the context of a "clean" R session. The goal is a good one, but the method used to attain it is not very effective. The only thing that `rm(list = ls())` does is remove any variables currently stored in the global environment, but this is only one of many different ways in which previously-executed code can alter the state of the R session. This approach doesn't provide any degree of safety from any of the following:

- packages that may have been loaded with `library()` and change which functions get executed by your code
- data sets or other environments that may have been added to the search path with `attach()` and alter the variables and functions that are visible to your code
- options that may have been set with `options()` and can alter how your code is interpreted 
- "hidden" variables in the global environment like `.Random.seed` that affect R code execution and are ignored by `rm(list = ls())`

This is by no means an exhaustive list, but it's enough to illustrate the many different factors that can can affect how your script executes: `rm(list = ls())` does not protect you against any of them. Jenny Bryan provides a good discussion of the [problems with the "object nuking" approach](https://www.tidyverse.org/blog/2017/12/workflow-vs-script/) on the tidyverse blog.

Because of this, a better practice is to **restart the R session** immediately before running the script. By running the script in a fresh R session, you're much less likely to encounter these issues.

## How does sessioncheck help?

The principle that underpins **sessioncheck** is that scripts should not attempt a "clean up" of the environment. The "state" of an R session is complicated, and so if your script detects evidence that the pre-existing R state isn't "clean", it should notify the human user that there might be a problem. It is the *users* responsibility to ensure that the session is clean. A script doesn't have the power to make that happen, and it shouldn't try. The best it can possibly do is call your attention to the issue.

If you accept this as the core principle, then the thing you want to do -- within that little snippet of code you insert at the top of the script -- is to have your script look at the state of the R session, and look for signs that it isn't being run in a clean environment. If it detects any such signs, it should warn the user, or (in cases where strict controls are required) produce an error that prevents the rest of the script from running. The purpose of **sessioncheck** is to provide tools that can do this. They are -- necessarily! -- based on heuristic rules that scan the session for signs that there might be an issue, so it will not be perfect. But at a bare minimum it can be guaranteed that this approach is strictly superior to using `rm(list = ls())`.

The core idea is very simple. In the typical case, you wouldn't load the package with `library()`. Instead, you would put this as the very line of your R script:

```{r}
#| label: basic-use
sessioncheck::sessioncheck("error")
```

When executed in a clean R session this function does nothing, and the script continues to run as it normally would. However, if the session is "contaminated" -- for some specific interpretation of what that means to call the session "contaminated" -- it will produce an error, prevent the script from running, and prompt the human user to consider restarting the R session. 

To illustrate what happens in a contaminated R session, I'll do a few things that most R users would agree fit the definition of a "contaminated" R environment: creating objects in the global environment, and attaching packages and environments to the search path.

```{r}
library(sessioncheck)                     # trigger check_attached_packages
my_data = data.frame(T = FALSE, F = TRUE) # trigger check_globalenv_objects
attach(my_data)                           # trigger check_attached_environments
```

These actions vary considerably in their riskiness. For the example above, here's a quick risk assessment:

- **Attaching packages**. It's possible that attaching the **sessioncheck** package with `library()` might cause problems, but the risk varies a lot depending on what the package does. I've tried to design **sessioncheck** safely, so most likely this specific command won't be a problem. But that won't always be true, and therefore your script should notify you about any unexpected packages that it detects. In any case, for the most part you wouldn't bother attaching this package. The typical use case for this package is to call `sessioncheck::sessioncheck()` once, at the top of the script. In the normal course of events you wouldn't be writing `library(sessioncheck)` in your script, so yes, the user should probably be informed. It's just politeness!

- **Adding objects to the global environment**. Adding the `my_data` data frame as a variable in the global environment is not in itself a bad thing, but in practice when you're running the same script over and over in an interactive context -- which is a thing that data analysts have to do a lot, much more so than software engineers -- you can quickly build up a lot of "detritus", bits and pieces and objects left behind from code you tried out previously. Over time it becomes a big problem, because it's harder for you to reason about what's going on in your data. The user should definitely be informed, because there are real risks here.

- **Attaching a (malicious!) environment to the search path**. Attaching data sets using `attach()` has been considered an extremely high-risk behavior in R for a long time. It creates chaos in the R session and the potential for horrific outcomes is very high. The example I constructed above is deliberately malicious, in an attempt to highlight just how bad `attach()` can be. If your script executes in an environment when `attach(my_data)` has previously been executed, you can no longer rely on R to make sensible judgements about arithmetic:

  ```{r}
  (2 + 2 == 5) == T
  ```

  On the face of it, R now believes that `2 + 2 = 5`. It doesn't actually believe this. The result happens because the prior call to `attach()` ensures that you can no longer rely on the shorthand that assumes `T == TRUE`. Indeed:

  ```{r}
  T == TRUE
  ```

  and in fact:

  ```{r}
  T
  ```

  In an ideal world an R script should never be using `T` as a shorthand for `TRUE` because it exposes the code to exactly this problem, but unfortunately there is a *massive* amount of old R code in current use that relies on the shortcut. The risk is real.

Noting that the R session is now contaminated in three ways, we can see what happens when `sessioncheck()` is called in a contaminated environment:

```{r}
#| error: true
sessioncheck::sessioncheck("error")
```

If this code were placed at the top of the user script, that script would simply refuse to run in this contaminated environment. It would produce an error that indicates which checks were performed, and lists specific issues that were detected in each test.

By way of comparison, consider what would happen if the traditional `rm(list=ls())` approach were employed. Executing this code will fix only one of the three problems:

```{r}
rm(list = ls())
```

The global environment is now clean, but any code that relies on `T` to mean `TRUE` will now break, because the `rm()` command has *not* detached the malicious environment:

```{r}
T
```

Even though `rm(list = ls())` has been invoked, *the R session is not safe*, and it will corrupt the behavior of any script that executes in this session. Importantly, notice that because `sessioncheck()` is stricter than the `rm()` approach, it still produces an error:

```{r}
#| error: true
sessioncheck::sessioncheck("error")
```

Under normal circumstances the thing that the user -- a human being -- would likely do next is restart the R session, and all of these problems would go away. That is the intended way to use **sessioncheck**. It is designed primarily for analysts who design their code in a human-in-the-loop fashion. It is explicitly **not** recommended for production code or any other situation where R code is executed without direct human oversight. 

Nevertheless, because the author of this vignette is also the person who created the contaminated R session, she also knows how to reverse it in an automated fashion:

```{r}
detach("my_data")
detach("package:sessioncheck")
sessioncheck::sessioncheck("error")
```

In general though, the job of cleaning the session or restarting R should be reserved for the human analyst. Tools such as **sessioncheck** should not attempt to do so.

